---
title: "ecoCopula - Graphical modelling and ordination with copulas"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{my-vignette}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r  include=FALSE}
knitr::opts_chunk$set(
  fig.width=6, fig.height=6,
  collapse = TRUE,
  comment = "#>"
)
```


### Overview

The ecoCopula package allows you to visualise multivariate discrete data with graphical models and ordination. The package was designed primarily for multivariate abundance data in ecology, however it can be applied to any multivariate discrete data. 


The two main functions are:

1. Copula graphical models (`cgr`) allow you to plot a *graph* which distinguishes between direct and indirect relationships between variables (species).
2. Copula ordination (`cord`) allows you to visualise how samples (sites) and variables (species) are located along several latent variables (unobserved environmental gradient).

To install ecoCopula is unfortunately a little annoying due to the mvabund package, on which it relies, throwing a warning, which gets converted to an error. This code generally works. If you have trouble please [email me](g.popovic@unsw.edu.au) for assistance.

### Installation

```{r eval=FALSE, include=TRUE}
# Sys.setenv(R_REMOTES_NO_ERRORS_FROM_WARNINGS = TRUE)
# devtools::install_github("gordy2x/ecoCopula",upgrade = "always")
# Sys.setenv(R_REMOTES_NO_ERRORS_FROM_WARNINGS = FALSE)
```

### Inside the black box 



### Bryce Canyon Vegetation Data

Estimates of cover class for all non-tree vascular plant species in 160 375m^2 circular sample plots. Species codes are first three letters of genus + first three letters of specific epithet.

```{r}
rm(list=ls())

library(labdsv)
data(brycesite)
X=brycesite
X$plotcode=substr(X$plotcode,3,5)
data(bryceveg)
old <- c(0.2,0.5,1.0,2.0,3.0,4.0,5.0,6.0)
new <- c(0.2,0.5,3.0,15.0,37.5,62.5,85.0,97.5)
bryceveg <- abundtrans(bryceveg,old,new)
bryce<- bryceveg[,-which(colSums(bryceveg>0) <= 20)]
```


```{r}
library(corrplot)
corrplot(cor(bryce),diag = FALSE,tl.cex=0.7)
```

```{r message=FALSE, warning=FALSE}
corrplot(cor(t(bryce)),diag = FALSE,tl.cex=0.3)
```

Two problems
1. Ordinal, shouldn't really use correlation - should model data as ordinal instead 
2. It is hard to see any patterns, especially when there are many species/sites.

### Modelling ordinal data
```{r}
library(ordinal)
library(mvabund)
source('C:/Users/gordy/Dropbox/Work/1. Projects/saint/manyanynew.R')
for(iVar in 1:dim(bryce)[2])
  bryce[,iVar]=factor(bryce[,iVar])
#Now fit a model using the clm function:
bryce_mod=Manyany("clm",bryce,abund~1,data=bryce) # usually manyany but has bug for R 4.0.0, will be fixed
plot(bryce_mod)
```

Residuals looks okay, so we know the model we fit is reasonable. Now let's look at ordinations using this model.

```{r}
library(ecoCopula)
ord_cord<-cord(bryce_mod)
plot(ord_cord,biplot=TRUE)
```

```{r}
sites=ifelse(X$slope>7,"red","black")
plot(ord_cord,site.col = sites)
```


```{r}
site_res <- data.frame(ord_cord$scores,X)
sp_res <- data.frame(ord_cord$loadings,species = colnames(bryce),genus = substr(colnames(bryce),1,3))
```

```{r}
library(ggplot2)
ggplot()+
  geom_text(aes(x=Factor1,y=Factor2,color = elev,label =plotcode ),site_res)
```

```{r}
ggplot()+
  geom_segment(aes(x=0,y=0, xend = Factor1, yend = Factor2,color = ),sp_res)+
  geom_text(aes(x = Factor1, y = Factor2,label = species,color=genus),data=sp_res)
```

```{r}
library(RColorBrewer)
alpha= 2.5
ggplot()+
  geom_point(aes(x=Factor1,y=Factor2,color = elev ),site_res)+
  geom_text(aes(x = Factor1*alpha, y = Factor2*alpha,label = species),data=sp_res)+
  scale_color_gradientn(colours = brewer.pal(n = 10, name = "PuOr"))+
  theme_classic()

```





### Graphical models

```{r}
ord_graph<-cgr(bryce_mod)
```


```{r}
plot(ord_graph, pad=1)
```

```{r}
igraph_out<-ord_graph$best_graph$igraph_out
# set_vertex_attr and ?set_edge_attr to add attributes for plotting

plot(igraph_out)
```


```{r}
library(tidyr)
library(tidygraph)
library(ggraph)


igraph_out %>% ggraph('fr') + # see ?layout_tbl_graph_igraph
    geom_edge_fan0(aes( colour = partcor, width=partcor)) +
    scale_edge_width(range = c(0.5, 3))+
    scale_edge_color_gradient2(low="#b2182b",mid="white",high="#2166ac")+
    geom_node_text(aes(label=name), repel = TRUE)+
    geom_node_point(aes(size=1.3))+
    theme_void() +
    theme(legend.position = 'none')

```





### Copula graphical models (`cgr`)


We will use the spider data (from the mvabund) as an example. It has counts of 12 hunting spiders at 28 sites, collected using pit traps.
```{r}
data(spider)
abund <- mvabund(spider$abund)
head(abund)
```

There are also some environmental variables.

```{r}
X <- as.data.frame(spider$x)
head(X)
```

```{r}
nsp=ncol(spider$abund)
dat=data.frame(spider$abund,spider$x)

dat%>% 
  pivot_longer(cols=1:nsp,names_to = "Species", values_to = "abund") %>% 
  mutate(some_bare_sand = ifelse(bare.sand>0,"yes","no")) %>% 
  ggplot(aes(y=Species,x=abund+1,fill=some_bare_sand))+ 
    geom_boxplot()+
    scale_x_log10()+
    theme_classic()
    
```

#presence-absence
```{r}
nsp=ncol(spider$abund)
dat_pa=data.frame(1*(spider$abund>0),spider$x)

dat_pa%>% 
  pivot_longer(cols=1:nsp,names_to = "Species", values_to = "abund") %>% 
  mutate(some_bare_sand = ifelse(bare.sand>0,"yes","no")) %>% 
  ggplot(aes(x=Species,fill=some_bare_sand))+ 
    facet_grid(~factor(abund), scales="free")+
    geom_bar()+
    coord_flip()+
    theme_classic()+
    theme(legend.position = "bottom")
    
    
```


To fit a graphical model, we first fit a model in the mvabund package, and then use the `cgr` function to extract the best *graph*. Here we will first fit a model with no predictors.

We might be interested in how these species are related. One way to do this is to look at their correlation. 

```{r}
library(corrplot)
corrplot(cor(abund),diag = FALSE)
```

```{r}
library(corrplot)
corrplot(cor(t(abund)),diag = FALSE)
```


### Copula ordination (`cord`)


```{r}
spid_lv=cord(spider_mod)
sites=ifelse(X$bare.sand>0,"red","black")
plot(spid_lv,biplot = TRUE,site.col = sites)
# corrplot(spid_lv$sigma, type = "upper",tl.col = "black",diag=FALSE,
#          method="shade",shade.col="gray")
```



We know how to deal prolerly with multivariate abundance data, we need to use glms, which for many species is easy to do with mvabund. 

```{r}
spider_mod=manyglm(abund~1)
# corrplot(cor(residuals(spider_mod)),diag = FALSE)
spid_graph=cgr(spider_mod)
plot(spid_graph,pad=0)
```


```{r}

library(corrplot)
corrplot(spid_graph$raw$cov)

corrplot(spid_graph$raw$cov, type = "upper",tl.col = "black",diag=FALSE,
         method="shade",shade.col="gray")

corrplot(spid_graph$best_graph$part, type = "upper",tl.col = "black",diag=FALSE,
         method="shade",shade.col="gray")

corrplot(spid_graph$best_graph$cov, type = "upper",tl.col = "black",diag=FALSE,
         method="shade",shade.col="gray")
```


```{r}
data(tikus)
tabund<- mvabund(tikus$abund)
#remove sepcies with fewer than 6 presences
tabund<-tabund[,-which(colSums(tabund>0) <= 8)]
corrplot(cor(tabund),diag = FALSE)


```


```{r}
tikus_mod=manyglm(tabund~1)
tik_graph=cgr(tikus_mod)

```

```{r}
plot(tik_graph,pad=1)
```
```{r}
tik_ord=cord(tikus_mod)
plot(tik_ord,biplot=TRUE)
plot(tik_ord,biplot=TRUE,site.col="red")

```



```{r}
library(ordinal)
library(labdsv)
data(bryceveg)

bryce<- bryceveg[,-which(colSums(bryceveg>0) <= 20)]
X=matrix(1,nrow(bryce),1)
for(iVar in 1:dim(bryce)[2])
  bryce[,iVar]=factor(bryce[,iVar])
#Now fit a model using the clm function:
bryce_mod=manyanynew("clm",bryce,abund~1,data=X)
plot(bryce_mod)
```

```{r}
ord_cord<-cord(bryce_mod)
plot(ord_cord,biplot=TRUE)
```

```{r}
ord_graph<-cgr(bryce_mod)
plot(ord_graph)
```


routes_tidy <- tbl_graph(nodes = vertex, edges = longedge, directed = FALSE)


# longedge$weight=1
A=network::network(longedge,vertex.attr = vertex, matrix.type = "edgelist", edge.attr = longedge)

plot(A,usearrows=FALSE)
