---
title: "ecoCopula - Graphical modelling and ordination with copulas"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{my-vignette}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```


### Overview

The ecoCopula package alows you to visualise multivariate discrete data with grapical models and ordination. The package was designed primarily for multivariate abindance data in ecology, however it can be applied to any multivatiate discrete data. 


The two main functions are:

1. Copula graphical models (`cgr`) allow you to plot a *graph* which distinguishes between direct and indeirect relationships between variables (species).
2. Copula ordination (`cord`) allows you to visualise how samples (sites) and variables (species) are located along several latent variables (unobserved environmental gradient).

To install ecoCopula is unfortunately a little annoying due to the mvabund package, on which it relies, throwing a warning, but this code generally works. If you have trouble please [email me](g.popovic@unsw.edu.au) for assistance.

### Installation

```{r eval=FALSE, include=TRUE}
Sys.setenv(R_REMOTES_NO_ERRORS_FROM_WARNINGS = TRUE)
devtools::install_github("gordy2x/ecoCopula",dependencies = FALSE)
Sys.setenv(R_REMOTES_NO_ERRORS_FROM_WARNINGS = FALSE)
```

### Inside the black box 


### Copula graphical models (`cgr`)
```{r setup}
library(ecoCopula)
```

### Copula ordination (`cord`)